# Introduction {#intro}

The heart and soul of multivariate statistics is **relationships between variables**.


How do we **describe** and **display** them?


How do we **measure** and **estimate** them?


How do we **test** them?


How do we **account for** them?


## Example: Paired t-tests {-}

Recall from your introductory Statistics class:

* **two-sample t-tests** compare the means of two **independent** samples
* a **paired** t-test compares the mean difference from two **dependent** samples


Consider

* Midterm grades in Section 01 versus midterm grades in Section 02
* Midterm grades in Stat 419 versus final grades in Stat 419


```{r, include = FALSE}
library(tidyverse)
library(kableExtra)

grades <- data.frame(
  names = c("Mario", "Luigi", "Wario", "Waluigi", "Bowser"),
  midterm_grades = c(85, 90, 65, 72, 41),
  final_grades = c(89, 93, 61, 78, 47)
)
```

```{r, echo = FALSE}
grades %>% kable()
```


```{r}
with(grades, 
  t.test(midterm_grades, final_grades, 
         alternative = "less", paired = FALSE)
)
```


```{r}
with(grades, 
  t.test(midterm_grades, final_grades, 
         alternative = "less", paired = TRUE)
)
```

This time, we accounted for the *relationship* between an individual's two exam scores!


```{r}
grades %>% 
  summarize_at(
    vars(midterm_grades, final_grades),
    funs(sd, mean))
```

The **variance** of the **difference of means** is:

$$\text{var}(\bar{X}_2 - \bar{X}_1) = (\sigma_1^2 + \sigma_2^2)/n$$

```{r}
var <- (19.3^2 + 19.4^2)/5
var
```

```{r}
sqrt(var)
```

But... they are not independent!

```{r}
grades %>% select(midterm_grades, final_grades) %>% cov()
```

The equation is actually:

$$\text{var}(\bar{X}_2 - \bar{X}_1) = (\sigma_1^2 + \sigma_2^2 - 2 \, \sigma_{12})/n$$

```{r}
actual_var <- (19.3^2 + 19.4^2 - 2*365)/5
actual_var
sqrt(actual_var)
```


## Goals of this class {-}

1. Expand your basic stats knowledge into the magical world of multiple variables.

2. Teach you to responsibly consider and test variable relationships.

3. Show you classic and cutting edge methods for multivariate data.

4. Have some fun with interesting datasets!



## Just for fun: My research {-}

- Data from **last.fm** music streaming site.

- **Samples:**  Listeners $(n = 1893)$

- **Variables:**  Artists $(d = 17,632)$


- **Goal:** Given an artist, recommend related artists based on listener data.

- Many variables (whether or not a listener listened to each artist), very dependent!


### Result 1: Paul McCartney {-}

|**Artist** | **Top 5 Tags** |
|---------|-------|
|Paul McCartney | sad, classic rock, cool, british, beautiful  |
 | The Beatles | 60s, classic rock, british, psychedelic, <3 |
|  George Harrison | classic rock, 70s, singer-songwriter, sad, british |
|  John Lennon | classic rock, singer-songwriter, 70s, british, male vocalists |



### Result 2:  Hannah Montana {-}

|**Artist** | **Top 5 Tags** |
|---------|-------|
| Hannah Montana | love at first listen, pop rock, soundtrack, amazing, female vocalist | 
|  Miley Cyrus | <3, catchy, love at first listen, amazing, pop rock | 
|  Rihanna | rnb, ballad, sexy, love, dance | 
|  Katy Perry | pop rock, <3, catchy, love, love at first listen | 
|  Britney Spears | catchy, female, sexy, amazing, dance | 
|  Ke$ha | love at first listen, dance, <3, pop, catchy | 
|  Lady Gaga | dance, female vocalist, love at first listen, catchy, sexy | 
|  Demi Lovato | love at first listen, <3, pop rock, catchy, female vocalist | 
|  Avril Lavigne | pop rock, canadian, pop punk, female, love at first listen | 
|  Taylor Swift | country, <3, catchy, love, amazing | 
|  Selena Gomez | <3, pop rock, love at first listen, catchy, love | 
|  Ashley Tisdale | <3, catchy, pop rock, ballad, awesome | 
|  Hilary Duff | favorites, amazing, sexy, pop rock, dance | 
|  Christina Aguilera | ballad, sexy, soul, rnb, amazing | 
|  Jonas Brothers | pop rock, <3, love, love at first listen, amazing | 
|  Beyonce | rnb, sexy, soul, ballad, female vocalist | 
|  Glee Cast | cover, love at first listen, love, catchy, soundtrack | 
