# Characterizing and Summarizing Multivariate Data

<script>
function example(ex_name) {
    var x = document.getElementById(ex_name);
    if (x.style.display === "none") {
        x.style.display = "block";
    } else {
        x.style.display = "none";
    }
}
</script>

<style>
.example {
  background-color: cornsilk;
  color: blue;
} 
.q {
  background-color: cornsilk;
  color: darkred;
  text-align: right;
} 
.alert {
  background-color: light red;
  color: dark blue;
  font: bold;
  text-align: center;
} 
button{
  margin:auto;
  display:block;
  float: right;
}
</style>

```{r}
library(knitr)
source("/Users/kbodwin/Dropbox/Teaching/kb_sweave_extras.R")
```



## Linear Combinations of Random Variables {-}


Suppose we have some constants (scalars) $a_1,a_2,\ldots,a_p$, and some variables
$y_1,y_2,\ldots,y_p$.  

A **linear
combination** of the $p$ elements of ${\bf y}$, denoted $z$ is
given by:

$$z=a_1y_1+a_2y_2+\cdots+a_py_p \, = \, {\bf a}'{\bf y}$$

<button onclick="example('z_scalar')"> !!!</button>
<div id="z_scalar" class="alert" style="display:none">
z is a scalar (or univariate random variable)
</br>
</div>

For example, suppose we
have a sample of 5 students' scores on 8 assignments - 5 homeworks, 2 midterms, and a final.

The $5 \times 8$ data matrix is given by:

```{r, echo = FALSE}
scores <- data.frame(
  HW1 = c(80, 75, 80, 70, 55),
  HW2 = c(90, 95, 80, 60, 80),
  HW3 = c(85, 80, 75, 100, 75),
  HW4 = c(70, 80, 70, 80, 65),
  HW5 = c(100, 90, 100, 95, 90),
  M1 = c(65, 75, 90, 80, 70),
  M2 = c(65, 60, 75, 90, 80),
  FIN = c(75, 70, 85, 90, 75)
)
```

```{r, echo = FALSE}
kable(scores)
```

$${\bf Y} = \left( \begin{array} `r rmat(scores)`\end{array} \right)$$

Suppose we consider the grading scheme that gives equal weight to each assignment:

$$z=\frac{1}{8}(y_1+y_2+y_3+y_4+y_5+y_6+y_7+y_8)$$

Then, Student 1 has a final grade of


```{r}
grade_1 <- sum(scores[1,])/8
grade_1
```

$$z=\frac{1}{8}(80 + 90 + 85 + 70 + 100 + 65 + 65 + 75) = `r grade_1`$$

If we wanted to calculate this grade average for all students simultaneously, we could represent this as a matrix calculation:

$$\overline{{\bf z}} = {\bf j}'{\bf Y}$$
```{r}
rowMeans(scores)
```


### Variances of linear combinations {-}

Recall that the **covariance** of a data matrix ${\bf Y}$ shows you the variances and covariances between all the variables:

```{r}
cov(scores)
```
</br>
<button onclick="example('ex_cov')"> Do students who do better on M1 and M2 tend to do better on the Final? </button>
<div id="ex_cov" class="example" style="display:none">
Yes!  The covariances are positive - 51.25 and 80.
</br>
</div>


Now, we'd like to know what the variance is average grades is; i.e., the variance of the *univariate* variable $z$.
</br>
<button onclick="example('ex_var')"> What is wrong with the approach below? </button>
<div id="ex_var" class="example" style="display:none">
We didn't account for the relationships between the variables!!!
</br>
</div>


```{r}
zs <- rowMeans(scores)
var(zs)
```

Instead, we need to use our matrix representation to think about the variance.

In univariate statistics, when you multiply a *variable* by a *constant*, you have to square the constant and multiply by the variance:

$$\text{var}(aX) = a^2 \text{var}(X)$$

The same idea applies in mulivariate statistics:  when you *matrix multiply* a *data matrix* by a *constant vector*, you have to "square" the vector and multiply by the covariance.

$$ \text{var}(z) = \text{var}({\bf Y}{\bf j}) \, = \, {\bf j}' \text{Cov}(Y) {\bf j}$$
$$ s^2(z) \, = \, {\bf j}' S_Y {\bf j}$$


<button onclick="example('q_var')"> ??? </button>
<div id="q_var" class="q" style="display:none">
This is just a math rule - which you might derive in other classes!
</div>

<button onclick="example('ex_var')"> What is the dimension of var(z)? </button>
<div id="ex_var" class="example" style="display:none">
1x1, it is a scalar!
</div>

## Practice {-}
Now consider the two weighting schemes for final grades:  First, the "average all scores" approach we saw above.  Second, a scheme where homework is worth 20%, midterms are worth 30%, and the final is worth 50%.

$$z^{(1)}=\frac{1}{8}(y_1+y_2+y_3+y_4+y_5+y_6+y_7+y_8)$$
$$z^{(2)}=\frac{1}{5}(y_1+y_2+y_3+y_4+y_5)\cdot0.2+\frac{1}{2}(y_6+y_7)\cdot0.3+y_8\cdot 0.5$$

We can write this in matrix form as

$$ z_2 = {\bf a}'{\bf y}$$
where ${\bf a}$ is...

<button onclick="example('ex_a')"> Click for answer </button>
<div id="ex_a" class="example" style="display:none">
a = (0.04, 0.04, 0.04, 0.04, 0.04, 0.15, 0.15, 0.5)
</div>

Thus, we can still use our matrix tricks - with ${\bf a}$ instead of ${\bf j}$ - to calculate everyone's scores, as well as the covariances!


$${\bf z} = {\bf Y}{\bf a}$$
$$s^2_z = {\bf a}' S_Y {\bf a}$$



```{r}
a <- c(0.04, 0.04, 0.04, 0.04, 0.04, 0.15, 0.15, 0.5)
Y <- as.matrix(scores)

all_grades <- Y %*% a
all_grades

var_grades <- t(a) %*% var(Y) %*% a
var_grades
sqrt(var_grades)
```
